On the reality of the quantum state once again

### 2022-01-17 Articles/2021-HarriganSpekkens/HarriganSpekkens.tex
HarriganSpekkens: another pass, added caption and tentative conclusions

\begin{abstract}
The ontological model framework introduced by Harrigan and Spekkens has been widely used to categorize quantum interpretations and as a basis for influential results such as the PBR theorem. In this paper we show that the model has a fundamental flaw: the epistemic structure it implicitly assumes does not follow the one dictated by quantum mechanics. Namely, the map between the epistemic states of the model and quantum density matrices is not even an order homomorphism in terms of the information entropy, consequently the epistemic content of mixed states is not mapped in a meaningful way. Moreover, by recasting the definitions of the model in terms of density matrices, we find that only $\psi$-epistemic interpretations are allowed by quantum mechanics, a result that is the opposite of the PBR theorem. We show that the cause of this mismatch is the effect of non-contextuality over mixed preparations that does not allow for a single decomposition of mixed states over pure states.
\end{abstract}

@@

Such a theorem had a remarkable resonance (\cite{Leifer:2014, Leifer:2014b}, \cite{Renner:2012}, \cite{Colbeck:2017}, \cite{Hardy:2013}, \cite{Maroney:2014}, \cite{Patra:2013}, \cite{Mansfield:2016}, \cite{Schlosshauer:2012, Schlosshauer:2013, Schlosshauer:2014}, \cite{Aaronson:2013}), and questions about its actual meaning are still discussed today: on the one hand, it rules out interpretations of QM where $\psi$ merely represents information. On the other, scholars recently shown that non-trivial epistemic as well as statistical approaches to QM are not refuted by the PBR argument (\cite{Ben:2017}, \cite{Rizzi:2018}, \cite{Oldofredi:2021}, \cite{DeBrota:2019}).

In this essay we are not going to rebut the theorem itself. We instead aim to carefully re-examine one of its premises, i.e.\ Harrigan and Spekkens (HS) classification between $\psi-$ontic and $\psi-$epistemic ontological models  \cite{Harrigan:2010} on top of which the PBR result is derived. We believe that a careful examination of this assumption forces us to re-evaluate the significance of the PBR conclusion.

HS proposed a rigorous classification in order to categorize the nature of the quantum state, i.e.\ to establish whether in a certain model $\psi$ corresponds to a real property of a quantum object, in that case the model is called $\psi-$ontic, or to some observer information, making it $\psi-$epistemic. While the original aim of this classification was to clarify Einstein's view of quantum theory, HS framework has been widely employed in the literature not only to categorize different interpretations, but also to argue what types of interpretations are admissible (\cite{Leifer:2013}, \cite{Leifer:2017}, \cite{Branciard:2014} \cite{Hermens:2021}, \cite{Wood:2015}, \cite{Ringbauer:2015}, \cite{Mazurek:2016}, \cite{Bartlett:2012}; cf.\ \cite{Oldofredi:2020b}, \cite{Lewis:2012}, \cite{Ladyman:2021} for critical discussions). Given the influence that HS framework and the PBR theorem have in the quantum foundations, it is worthwhile to look again at the former with more scrutiny in order to understand whether it is a sound basis to draw conclusions on the nature of the quantum state.

In this paper we will show two things:
\begin{enumerate}
	\item the information entropy as calculated on epistemic states does not match the one provided by quantum mechanics, meaning that the epistemic structures provided by QM and HS are incompatible;
	\item the density matrices associated with pure states overlap for non-orthogonal states (i.e. $\rho_\psi \rho_\phi \neq 0$), meaning that $\psi$-ontic models are ruled out, contrary to the PBR claim.
\end{enumerate}
Hence, our conclusion is that HS distinction is fundamentally flawed. The main problem lies in assuming that the epistemic state identifies a single probability distribution over the whole set of ontic states. We show that this, in essence, is an assumption of non-contextuality, which is also inherited by the PBR theorem, casting doubts on its validity.

@@

We briefly review the main features of the classification provided by HS starting with the usual operational setting for QM. We have a preparation protocol $P$, which will be associated with a density operator $\rho$ on the relevant Hilbert space, and a measurement protocol $M$, represented by a POVM $\{ E_k\}$ where each $k$ represents a possible measurement outcome. The probability of obtaining a particular $k$ given a particular preparation $P$ and measurement $M$ is given by the generalized Born rule
\begin{equation}
	p(k|M, P)=\textrm{tr}(\rho E_k).
\end{equation}

@@

Let us consider more closely the space of epistemic states given by $p(\lambda | P)$ and equation \ref{epistemic_mixing}. To make the discussion more concrete, we assume to have a $\psi$-complete model of a single qubit representing the direction of spin.

@@

\begin{figure}
\includegraphics[scale=.4]{fig2}
\caption{\footnotesize{}}\label{fig_map}
\end{figure}

@@

It is widely understood that one feature of quantum theory is contextuality: a measurement cannot be conceived as revealing pre-existing values. We may be tempted to think that we have one context before and one after the measurement. This does not work because we have a related dual effect on preparations. In the same way that we cannot understand quantum measurements as selecting from a single joint probability distribution for all properties, we cannot understand quantum preparations as constructing a single joint probability distribution over all properties. To fully understand how this works, let us review how probability spaces behave in both classical and quantum mechanics.

@@

Suppose to consider two density matrices $\rho_1$ and $\rho_2$ with different eigenstates and then to create the mixture $\rho = p_1 \rho_1 + p_2 \rho_2$. The context of the mixture will not be the same as the original contexts. It will, in fact, even depend on $p_1$, $p_2$. That is, when mixing preparations we are not simply combining two probability distributions over the same context, like in classical mechanics, but we are combining the contexts themselves to create a new one. Contextuality not only affects measurements, but, dually, preparations as well. And this affects how information entropy is calculated and the overall epistemic structure.

To sum up, assuming that $p(\lambda|P)$ exists means assuming that there is a single probability space, a single $\sigma$-algebra, a single set of propositions to which all probabilities are assigned. This is exactly assuming non-contextuality. While this can be made to work for pure state-to-pure state transitions during measurements, mixing preparations leads to an epistemic structure that is fundamentally non-contextual, leading to inconsistencies with quantum information. To allow for contextuality we should assume that for each observable $O$ we have a probability $p_O(\lambda_O|P)$ of having prepared one of the eigenstates, such that $\sum_{\lambda_O} p_O(\lambda_O|P) = 1$. While we have a single state space $\Lambda$, we do not have a single probability space over the whole space. The entropy of the preparation would be the lowest entropy across all contexts.

@@

>>>>

\begin{abstract}
The Pusey, Barrett and Rudolph (PBR) theorem concludes that quantum states cannot be taken to represent mere information about the system. This result is based on the ontological model framework proposed by Harrigan and Spekkens (HS). In this paper we show that the HS framework has a fundamental problem: the epistemic structure it implicitly assumes does not follow the one dictated by quantum mechanics. Namely, the map between the epistemic states of the model and quantum density matrices preserves neither the value nor ordering of the information entropy. Consequently the epistemic content of mixed states is not mapped in a meaningful way. The problem stems from the assumption that an epistemic state is characterized by a \emph{single} probability measure, which is essentially an assumption of non-contextuality. Given this fundamental issue, every result that uses the HS framework, including the PBR theorem, should be carefully reexamined.
\end{abstract}

@@

Such a theorem had a remarkable resonance (\cite{Leifer:2014, Leifer:2014b, Renner:2012, Colbeck:2017, Hardy:2013, Maroney:2014, Patra:2013, Mansfield:2016, Schlosshauer:2012, Schlosshauer:2013, Schlosshauer:2014, Aaronson:2013}), and questions about its actual meaning are still discussed today: on the one hand, it rules out interpretations of QM where $\psi$ merely represents information. On the other, scholars recently showed that non-trivial epistemic as well as statistical approaches to QM are not refuted by the PBR argument (\cite{Ben:2017, Rizzi:2018, Oldofredi:2021, DeBrota:2019}).

In this paper we are not going to rebut the theorem itself. We instead aim to carefully reexamine one of its premises, i.e.\ Harrigan and Spekkens' (HS) classification between $\psi-$ontic and $\psi-$epistemic ontological models  \cite{Harrigan:2010} on top of which the PBR result is derived. We believe that a careful examination of this assumption forces us to reevaluate the significance of the PBR conclusion.

HS proposed a rigorous classification in order to categorize the nature of the quantum state, i.e.\ to establish whether in a certain model $\psi$ corresponds to a real property of a quantum object, in which case the model is called $\psi-$ontic, or to some observer information, making it $\psi-$epistemic. While the original aim of this classification was to clarify Einstein's view of quantum theory, the HS framework has been widely employed in the literature not only to categorize different interpretations, but also to argue what types of interpretations are admissible (\cite{Leifer:2013, Leifer:2017, Branciard:2014, Hermens:2021, Wood:2015, Ringbauer:2015, Mazurek:2016, Bartlett:2012}; cf.\ \cite{Oldofredi:2020b, Lewis:2012, Ladyman:2021} for critical discussions). Given the influence that the HS framework and the PBR theorem have in quantum foundations, it is worthwhile to look again at the former with more scrutiny in order to understand whether it is a sound basis from which to draw conclusions on the nature of the quantum state.

In this paper we will show two things:
\begin{enumerate}
	\item the information entropy as calculated on HS epistemic states does not match the one provided by quantum mechanics, therefore the epistemic structures provided by QM and HS are incompatible;
	\item in QM the density matrices associated with pure states overlap for non-orthogonal states (i.e. $\rho_\psi \rho_\phi \neq 0$), which rules out $\psi$-ontic models
\end{enumerate}
Combining these results with PBR, our conclusion is that the HS classification itself is fundamentally problematic. The main issue lies in assuming that the epistemic state identifies a \emph{single} probability distribution over the whole set of ontic states. We show that this, in essence, is an assumption of non-contextuality, which is also inherited by the PBR theorem, casting doubts on its validity.

@@

We briefly review the main features of the classification provided by HS starting with the usual operational setting for QM. We have a preparation protocol $P$, associated with a density operator $\rho$ on the relevant Hilbert space, and a measurement protocol $M$, represented by a POVM $\{ E_k\}$ where each $k$ represents a possible measurement outcome. The probability of obtaining a particular $k$ given a particular preparation $P$ and measurement $M$ is given by the generalized Born rule
\begin{equation}
	p(k|M, P)=\textrm{tr}(\rho E_k).
\end{equation}

@@

Let us consider more closely the space of epistemic states given by $p(\lambda | P)$ and equation \ref{epistemic_mixing}. To make the discussion more concrete, we assume we have a $\psi$-complete model of a single qubit representing the direction of spin.

@@

\begin{figure}
\includegraphics[scale=.4]{fig2}
\caption{\footnotesize{The function $\iota$ maps the epistemic states $E_{HS}$ of a $\psi$-ontic model of a single qubit to the respective density matrices $E_Q$. Note that neither the value nor the ordering of the information entropy is preserved.}}\label{fig_map}
\end{figure}

@@

It is widely understood that one feature of quantum theory is contextuality: a measurement cannot be conceived as revealing pre-existing values. We may be tempted to think that we have one context before and one after the measurement. This does not work because we have a related dual effect on preparations. In the same way that we cannot understand quantum measurements as selecting from a single joint probability distribution for all properties, we cannot understand quantum preparations as constructing a single joint probability distribution over all properties. To fully understand how this works, let us review how probability spaces are defined in both classical and quantum mechanics.

@@

Consider two density matrices $\rho_1$ and $\rho_2$ with different eigenstates and their mixture $\rho = p_1 \rho_1 + p_2 \rho_2$. The context of the mixture will not be the same as the original contexts. It will, in fact, even depend on $p_1$, $p_2$. That is, when mixing preparations we are not simply combining two probability distributions over the same context, like in classical mechanics, but we are combining the contexts themselves to create a new one. Contextuality not only affects measurements, but, dually, preparations as well. And this affects how information entropy is calculated and the overall epistemic structure.

To sum up, assuming $p(\lambda|P)$ exists means assuming there is a single probability space, a single $\sigma$-algebra, a single set of propositions to which all probabilities are assigned. This is exactly assuming non-contextuality. While this can be made to work for pure state-to-pure state transitions during measurements, mixing preparations leads to an epistemic structure that is fundamentally non-contextual, leading to inconsistencies with quantum information. To allow for contextuality we should assume that for each observable $O$ we have a probability $p_O(\lambda_O|P)$ of having prepared one of the eigenstates, such that $\sum_{\lambda_O} p_O(\lambda_O|P) = 1$. While we have a single state space $\Lambda$, we do not have a single probability space over the whole space. The entropy of the preparation would be the lowest entropy across all contexts.

@@

While work in quantum foundations typically focuses on measurements and their effect on pure states, equal importance must be given to preparations and their effect on mixed states. The different behavior of entropy in classical and quantum mechanics has profound implications not just for information theory, but also for statistical mechanics and its measurable predictions. A full solution to the ``measurement problem'' will therefore have to address the dual ``preparation problem.''
